### 2015-02-09 Studies/2012_PG_County/Report/PG_Co_Fire.tex
PG Co: update report from WERB comments

% PDF metadata
\hypersetup{
  pdftitle={Simulation of a Wind Driven Basement Fire - Riverdale Heights, MD},
  pdfauthor={Weinschenk, Craig G., Overholt, Kristopher J., and Madrzykowski, Daniel},
  pdfsubject={Insert abstract here.},
  pdfkeywords={insert; keywords; here}
}

@@

\headerB{
Simulation of a Wind Driven \\
Basement Fire - \\
Riverdale Heights, MD \\
}

@@

January 2015}}

@@

On February 24, 2012, a fire in residential structure injured seven Prince George's County firefighters. NIST examined the fire dynamics of this incident at the request of the Prince George's Country Fire/Emergency Medical Services (EMS) Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this single-story residential structure with a basement using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Prince George's Fire Department likely experienced during the course of their interior operations.
\item To provide a basis for discussing tactical operations.
\end{enumerate}

@@

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident was documented by the Safety Investigation Team of the Prince George's County Fire/EMS Department~\cite{PGCounty2013}. At 21:11 hours on February 24, 2012 Prince George's Fire/EMS Department responded to a structure fire in Riverdale Heights, Maryland with a fire in the basement~\cite{PGCounty2013}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The Prince George's County report~\cite{PGCounty2013} includes narratives of actions of the response personnel. Note that term following the apparatus number (e.g. Truck 809 Can) refers to the seating assignment and operational duties of the firefighters on board. The following narrative of the incident, focused on the events which led to serious burn injuries to two emergency-personnel, was taken from the Prince George's County report~\cite{PGCounty2013}:
\begin{quote}
Truck 809 responded from quarters, as the first due truck company, and arrived on 57th Avenue right behind Engine 807B. When Truck 809 stopped to let Engine 807B layout, Truck 809 Can dismounted the truck and had to walk up the street as it pulled away. Truck 809 Officer observed a lot of smoke moving extremely fast through the front yard and across the street upon exiting the truck. Truck 809 Officer and Truck 809 Forcible Entry proceeded directly to the front door \ldots of the structure. Both of them entered the structure with full PPE, including SCBA, but without the protection of a hose line. Truck 809 Officer and Truck 809 Forcible Entry began primary searches on the first floor. At some point during the primary search, as conditions worsened, the front door closed, trapping Truck 809 Officer and Truck 809 Forcible Entry inside the structure. This situation could not have occurred if a charged hose line was operating inside the structure at the time the door slammed shut. They were the only firefighters operating on the first floor at that time.

@@

\begin{table}[!h]
\centering
\captionof{table}[Abridged approximate fire event timeline]{Abridged Prince George's Fire/EMS approximate fire event timeline~\cite{PGCounty2013}.}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time              &  Fire Behavior / Fireground Operation  \\
{[hh:mm:ss]}               &                                        \\
\midrule
21:08:26   & First 911 call is made. \\[.25cm]
\multirow{3}{*}{21:11:03}  &   \multirow{3}{*}{\parbox{10cm} {Dispatch for a ``box'' alarm assignment- four engine companies, two truck companies, one rescue squad, and a commanding officer.}}    \\
                           &         \\
                           &         \\[.25cm]
21:12:55  &  First unit (Engine 807B) arrived on-scene.  \\[.25cm]
21:13:44  & Volunteer Chief 809A establishes incident command (IC).   \\[.25cm]
21:13:56  & Truck 809 arrived on-scene. \\[.25cm]
21:14:00  & Front door opened. \\[.25cm]
21:14:42  & Truck 809 enters structure. \\[.25cm]
21:15:47  & Basement door opened and water applied to fire. Front door blown closed. \\[.25cm]
21:15:51  & Lower pane of front egress window broke. \\[.25cm]
21:16:01  & Top pane of front egress window broke. \\[.25cm]
21:16:16  & Truck 809 officer bailout through window. \\[.25cm]
21:16:44  & Front door opened. \\[.25cm]
21:17:15  & Truck 809 forcible entry firefighter rescued. \\[.25cm]
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a one story, single-family structure with a basement. The structure was platform-framed wood construction with a gable roof. \footnote{Platform-framed construction means that floor joists are completely covered with a sub-floor. This forms a platform for which walls are built from.} The structure had a first floor footprint of 9.14~m (30~ft) by 7.77~m (25.5~ft) with the basement being slightly smaller, 9.14~m (30~ft) by 7.16~m (23.5~ft), as show in Fig.~\ref{fig:simp_geom}. This resulted in a first floor area of 71~m$^2$ (764~ft$^2$) and a basement floor area of 65~m$^2$ (700~ft$^2$). The interior walls were composed of gypsum board over wood framing~\cite{PGCounty2013}. There was a single flight of stairs in the center of the structure which connected the first floor to the basement.

@@

Note that using the input lines above will invoke the simple chemistry reaction mechanism in FDS in which fuel and air react to form only CO$_2$, CO, H$_2$O, soot, and N$_2$. If the inclusion for other combustion products is desired, then the user must explicitly define those species and the chemical reaction that produces them~\cite{FDS_Users_Guide}. Based on the above input lines, FDS uses the default, mixing-controlled fast chemistry combustion model. This mechanism states that the rate of fuel consumption is proportional to both the local limiting reactant concentration and the local rate of mixing and extinction is based on the critical flame temperature approach~\cite{FDS_Math_Guide}. While FDS provides users with the option to use a more complex finite-rate combustion mechanism, there is not sufficient evidence in this case to justify deviating from the default specifications.

@@

Initial simulations were conducted using the two fires totaling 5.3~MW, however, the fire behavior did not align with observations made during the fire incident or post-incident images. Specifically there was insufficient fuel to drive the basement to an underventilated condition. One reason is that while the structure was unoccupied at the time of the fire~\cite{PGCounty2013}, there were additional fuels that were not accounted for: transient fuels, sparse kitchen cabinetry, and an unknown bathroom vanity among other fuels. A second reason is that the 50~kW/m$^2$ HRRPUA used to estimate the fire size does not account for an observed increase in burning rate during wind driven fires~\cite{madrzykowski2009fire,kerber2009fire}. It is not clear how to quantify the impact of wind on burning rate within a structure from first principles, so several iterations of the model with increased fire sizes were run. The final estimated fire size was a 1.8~MW fire in the bathroom and a 6.75~MW fire in the kitchen. This resulted in a total peak specified fire size of 8.55~MW.

Since the time at which the fire started is not known relative to the notification of a fire or how fast the fire spread, several simplifications were made. The fires started at the beginning of the simulation and increased to a value of 8.55~MW in 10~s. The 10-s ramp up time was used to prevent abrupt changes in the velocity at the fire boundary condition in FDS. Since the actual ignition time is unknown, the fire was ramped quickly to reflect the conditions determined from reports and post-incident images. The second assumption is that the sources of the gas phase wood fuel have constant areas at fixed locations in the basement (1 source for the kitchen, 1 source for the bathroom).  At 207~s, using a simple decay model, the fire size decreases linearly to 25~\% of its peak value in 60~s. This is to account for the impact of firefighters putting water on the basement fire (Table~\ref{tab:fire_info}). The following lines define the source fires in the FDS input file:

@@

\section{Ventilation}
\label{Vents}
The simulation in this study accounts for changes in the ventilation due to a combination of fire department operations (opening doors), fire acting on the structure (fire breaching windows) and wind (restricting flow, closing doors). The ventilation times and ventilation areas represent the best understanding of the incident. The times of ventilation changes for doors and windows are provided in Table~\ref{tab:vents}.

During the time at which the firefighters exited the structure or were rescued from the structure, the state of some windows and doors in the structure was known. The state of the vents (opened versus closed) resulted in the establishment of a flow path in the interior stairwell leading from the basement to the first floor. The exact time of operation for other doors/windows was not known. Therefore, the following the two basement windows on the left rear corner of the structure were assumed to be open at the start of the simulation. The states of the front door and front window were specified at certain times based on observations (i.e. forcible entry, wind effects, forced window breakage) from post-incident reports.

Structure leakage has been shown to be important when modeling enclosure fires~\cite{beal2009}. The open rear basement windows at the start of the simulation are significantly larger than the total effective area of leakage of a structure of this type. Therefore, any leakage in the structure would have a negligible impact on the fluid mechanics within the structure and is not included in this study.

In addition to the state of openings within the structure, outlined in Table~\ref{tab:vents}, an 8.9~m/s (20~mph) wind velocity boundary condition was included. The flow direction of the wind was from the rear toward the front of the structure, perpendicular to the rear of the structure.  Based on local weather conditions, mean wind velocities ranged between 4.5~m/s (10~mph) and 8.9~m/s (20~mph) with gusts up to 13.4~m/s (30~mph) with a direction that angled toward the back left of the structure~\cite{PGCounty2013}.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{../Figures/pg_county_meshtotal2}
\caption[Overhead front-left side of the structure within the computational domain.]{Overhead front-left view of the structure within the computational domain (with 17 meshes). The entire computational domain is 24~m (79~ft) by 16~m (53~ft) by 8~m (26~ft).}
\label{fig:mult_mesh}
\end{figure}

@@

\begin{table}[!ht]
\centering
\captionof{table}{Relevant fire model input parameters}\label{tab:mod_param}
\begin{tabular}{lll}
\toprule[1.5pt]
Parameter                                                  & Description                                   & Discussion                            \\
\midrule
Simulation Time                                            &  5~min                                        & --                                    \\ [.25cm]
\multirow{2}{*}{Grid Cell Size}                            &  10~cm around house                           & \multirow{2}{*}{Section~\ref{mesh}}   \\ 
                                                           &  20~cm for wind                               &                                       \\ [.25cm]
Ambient Temperature$^*$                                    &  20~$^{\circ}$C (68~$^{\circ}$F)              & --                                    \\ [.1cm]
\multirow{4}{*}{Reaction: Wood~\cite{SFPE:Tewarson}}       &  Formula: $\C\H_{1.7}\O_{0.74}\N_{0.002}$     & \multirow{4}{*}{Section~\ref{fire}}   \\
                                                           &  CO Yield: 0.004~$\mathrm{kg}/\mathrm{kg}$    &                                       \\
                                                           &  Soot Yield: 0.015~$\mathrm{kg}/\mathrm{kg}$  &                                       \\
                                                           &  Heat of Combustion: 16,400~kJ/kg             &                                       \\ [.25cm]
\multirow{2}{*}{Peak HRR}                                  &  Basement Fire: 6.75~MW                       & \multirow{2}{*}{Section~\ref{fire}}   \\ 
                                                           &  Bathroom Fire: 1.8~MW                        &                                       \\ [.25cm]                     
\multirow{3}{*}{Material: Wood~\cite{Incropera:1} }        &  $k$: 0.12~\si{W/(m.K)})                      & \multirow{3}{*}{Section~\ref{matl}}   \\
                                                           &  $\rho$: 510~kg/m$^3$                         &                                       \\
                                                           &  $c_{p}$: 1.38~\si{kJ/(kg.K)}                 &                                       \\ [.25cm]
\multirow{3}{*}{Material: Gypsum Board~\cite{WAKILI2007}}  &  $k$: 0.28~\si{W/(m.K)}                       & \multirow{3}{*}{Section~\ref{matl}}   \\ 
                                                           &  $\rho$: 810~kg/m$^3$                         &                                       \\ 
                                                           &  $c_{p}$: 1.0~\si{kJ/(kg.K)}                  &                                       \\
\bottomrule[1.25pt]
\end{tabular}\par
\footnotesize
$^{*}$Initial interior temperatures were assumed to be 20~$^{\circ}$C; exterior temperatures were 10~$^{\circ}$C~\cite{PGCounty2013}.
\normalsize
\end{table}

@@

\begin{table}[!ht]
\caption[Fire incident and simulation event timeline]
{Fire incident and simulation event timeline}
\begin{tabular}{ccl}
\toprule
Incident Time  &  Simulation  &  Fire Behavior / Fireground Operation                                \\
               &  Time        &                                                                      \\
{(hh:mm:ss)}   &  {(s)}       &                                                                      \\
\midrule
21:08:26       &              &  First 911 call is made.                                             \\
               &              &                                                                      \\
21:11:03       &              &  Dispatch for a``box'' alarm assignment- four engine companies       \\
               &              &  two truck companies, one rescue squad, and a commanding officer.    \\
               &              &                                                                      \\
21:21:20       &  0           &  FDS simulation begins.                                              \\
               &              &                                                                      \\
21:12:55       &              &  First unit (Engine 807B) arrived on-scene.                          \\
               &              &                                                                      \\
21:13:44       &              &  Volunteer Chief 809A establishes incident command (IC).             \\
               &              &                                                                      \\
21:13:56       &              &  Truck 809 arrived on-scene.                                         \\
               &              &                                                                      \\
21:14:00       &  100         &  Front door opened.                                                  \\
               &              &                                                                      \\
21:14:42       &              &  Truck 809 enters structure.                                         \\
               &              &                                                                      \\
21:15:47       &  207         &  Front door closed over.                                             \\
               &              &  Basement door opened and water applied to fire.                     \\
               &              &                                                                      \\
21:15:51       &  211         &  Lower pane of front egress window broke.                            \\
               &              &                                                                      \\
21:16:01       &  221         &  Top pane of front egress window broke.                              \\
               &              &                                                                      \\
21:16:16       &              &  Truck 809 officer bailout through window.                           \\
               &              &                                                                      \\
21:16:47       &  267         &  Front door opened.                                                  \\
               &              &                                                                      \\
21:17:15       &              &  Truck 809 forcible entry firefighter rescued.                       \\
               &              &                                                                      \\
\bottomrule
\end{tabular}
\footnotesize
\\ For specific information regarding ventilation areas used in the simulation, refer to Table~\ref{tab:vents}.
\normalsize
\label{tab:combined_timeline}
\end{table}

@@

\section{Heat Release Rate}
\label{HRR}
In FDS, a HRR is specified, which results in a specified amount of fuel vapor (or pyrolyzate) being released from a fuel surface. FDS then determines the amount of combustion that occurs throughout the simulation based on the amount of available fuel and oxygen at a given location. As a result, the prescribed HRR that was input into the simulation can be different than the HRR calculated by FDS based on the ventilation conditions.

Figure~\ref{fig:PG_Total_HRR} shows the HRR in the structure vs. time based on the fires described in Section~\ref{fire}. In this figure, the solid line represents the prescribed HRR that was input into the simulation (based on a prescribed mass flux), and the dashed line represents the HRR inside the structure that was calculated by FDS based on the ventilation conditions. The vertical lines represent the times at which ventilation occurred: front door opening and closing as well as the breaking of a front window for emergency escape. See Table~\ref{tab:vents} for a detailed list of the ventilation that occurring during the incident. The maximum size of the fire specified in this study was selected such that the conditions aligned with observations and fuel loads from the actual fire incident.

\begin{figure}[!ht]
\includegraphics[width=5in]{../Figures/PG_Total_9MW_HRR}
\caption[Interior prescribed and calculated interior HRRs vs. time from the simulation.]
{Comparison of prescribed and calculated interior HRRs from the FDS simulation. The vertical lines indicate the times at which ventilation occurred.}
\label{fig:PG_Total_HRR}
\end{figure}

In Fig.~\ref{fig:PG_Total_HRR}, the prescribed HRR that was input into the simulation is different than the HRR calculated by FDS based on the ventilation conditions. From 0~s to approximately 9~s, the calculated HRR was in agreement with the prescribed HRR because there was an adequate amount of oxygen for all of the fuel to combust. The differences in prescribed versus predicted HRR is a result of either fuel accumulation within the structure not combusting because of a lack of oxygen or combustion occurring exterior to the structure where fuel and oxygen can mix and react. Fig.~\ref{fig:PG_9MW_HRR} shows the total calculated HRR within the computational domain (includes combustion occurring exterior to the structure) compared to the prescribed HRR curve.

@@

A comparison of the differences between the HRR inside the structure (Fig.~\ref{fig:PG_Total_HRR}) and the total HRR (Fig.~\ref{fig:PG_9MW_HRR}) over the first 100 seconds of the simulation indicated that on average there was the equivalent of an approximate 0.7 MW fire(s) occurring through open external vents (basement windows). This aligns with observed post-fire damage and post-incident report~\cite{PGCounty2013}. Fig.~\ref{fig:ext_fire_dam} shows a post-fire image of the back left corner of the structure as well as a Smokeview rendering of the fire from the FDS simulation prior to the front door being opening by firefighters.

@@

After 100 seconds of simulation time, 5~\sfrac{1}{2}~minutes after the first 911 call, firefighters open the front door to make entry to the first floor of the structure. The change in ventilation to the structure impacted the HRR. After the door was opened, combustion no longer tool place outside of the structure, however as shown in Fig.~\ref{fig:PG_Basement_9MW_HRR}, the basement HRR was lower than the total HRR within the domain.

\begin{figure}[!ht]
\includegraphics[width=5in]{../Figures/PG_Basement_9MW_HRR}
\caption[Prescribed, calculated domain, and calculated basement HRRs vs. time from the simulation.]
{Comparison of prescribed, total calculated domain HRR and basement calculated HRRs from the simulation. The vertical lines indicate the times at which ventilation occurred.}
\label{fig:PG_Basement_9MW_HRR}
\end{figure}

>>>>

% PDF metadata
\hypersetup{
  pdftitle={Simulation of a Wind Driven Residential Basement Fire - Riverdale Heights, MD},
  pdfauthor={Weinschenk, Craig G., Overholt, Kristopher J., and Madrzykowski, Daniel},
  pdfsubject={Insert abstract here.},
  pdfkeywords={insert; keywords; here}
}

@@

\headerB{
Simulation of a Residential \\
Wind Driven Basement Fire - \\
Riverdale Heights, MD \\
}

@@

February 2015}}

@@

On February 24, 2012, a fire in an uninhabitated residential structure injured seven Prince George's County firefighters. NIST examined the fire dynamics of this incident at the request of the Prince George's Country Fire/Emergency Medical Services (EMS) Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this single-story residential structure with a basement using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Prince George's Fire Department likely experienced during the course of their interior operations.
\item To provide a basis for analyzing tactical operations in terms of fire dynamics.
\end{enumerate}

@@

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident was documented by the Safety Investigation Team of the Prince George's County Fire/EMS Department~\cite{PGCounty2013}. At 21:11 hours on February 24, 2012 Prince George's Fire/EMS Department responded to a residential structure fire in Riverdale Heights, Maryland with a fire in the basement~\cite{PGCounty2013}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The Prince George's County report~\cite{PGCounty2013} includes narratives of actions of the response personnel. Note that term following the apparatus number (e.g. Truck 809 Can) refers to the seating assignment and operational duties of the firefighters on board. The following narrative of the incident, focused on the events which led to serious burn injuries to two emergency-personnel, was taken from the Prince George's County report~\cite{PGCounty2013}:
\begin{quote}
Truck 809 responded from quarters, as the first due truck company, and arrived on 57th Avenue right behind Engine 807B. When Truck 809 stopped to let Engine 807B layout, Truck 809 Can dismounted the truck and had to walk up the street as it pulled away. Truck 809 Officer observed a lot of smoke moving extremely fast through the front yard and across the street upon exiting the truck. Truck 809 Officer and Truck 809 Forcible Entry proceeded directly to the front door \ldots of the structure. Both of them entered the structure with full PPE, including SCBA, but without the protection of a hose line. Truck 809 Officer and Truck 809 Forcible Entry began primary searches on the first floor. At some point during the primary search, as conditions worsened, the front door closed, trapping Truck 809 Officer and Truck 809 Forcible Entry inside the structure. This situation could not have occurred if a charged hose line was operating inside the structure at the time the door slammed shut. They were the only firefighters operating on the first floor at that time.

@@

\begin{table}[!h]
\centering
\captionof{table}[Abridged approximate fire event timeline]{Abridged Prince George's Fire/EMS approximate fire event timeline~\cite{PGCounty2013}.}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time              &  Fire Behavior / Fireground Operation  \\
{[hh:mm:ss]}               &                                        \\
\midrule
21:08:26   & First 911 call is made. \\[.25cm]
\multirow{3}{*}{21:11:03}  &   \multirow{3}{*}{\parbox{10cm} {Dispatch for a ``box'' alarm assignment- four engine companies, two truck companies, one rescue squad, and a commanding officer.}}    \\
                           &         \\
                           &         \\[.25cm]
21:12:55  & First unit (Engine 807B) arrived on-scene.  \\[.25cm]
21:13:44  & Volunteer Chief 809A establishes incident command (IC).   \\[.25cm]
21:13:56  & Truck 809 arrived on-scene. \\[.25cm]
21:14:00  & Front door opened. \\[.25cm]
21:14:42  & Truck 809 personnel enter structure. \\[.25cm]
21:15:47  & Basement door opened and water applied to fire. Front door blown closed. \\[.25cm]
21:15:51  & Lower pane of front egress window broken. \\[.25cm]
21:16:01  & Top pane of front egress window broken. \\[.25cm]
21:16:16  & Truck 809 officer bailout through front window. \\[.25cm]
21:16:44  & Front door opened. \\[.25cm]
21:17:15  & Truck 809 forcible entry firefighter rescued. \\[.25cm]
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a one story, single-family structure with a basement. The structure was platform-framed wood construction with a gable roof. \footnote{Platform-framed construction means that floor joists are completely covered with a sub-floor. This forms a platform from which walls are built.} The structure had a first floor footprint of 9.14~m (30~ft) by 7.77~m (25.5~ft) with the basement being slightly smaller, 9.14~m (30~ft) by 7.16~m (23.5~ft), as show in Fig.~\ref{fig:simp_geom}. This resulted in a first floor area of 71~m$^2$ (764~ft$^2$) and a basement floor area of 65~m$^2$ (700~ft$^2$). The interior walls were composed of gypsum board over wood framing~\cite{PGCounty2013}. There was a single flight of stairs in the center of the structure which connected the first floor to the basement.

@@

Note that using the input lines above will invoke the simple chemistry reaction mechanism in FDS in which fuel and air react to form only CO$_2$, CO, H$_2$O, soot, and N$_2$. If the inclusion for other combustion products is desired, then the user must explicitly define those species and the chemical reaction that produces them~\cite{FDS_Users_Guide}. Based on the above input lines, FDS uses the default, mixing-controlled fast chemistry combustion model. This mechanism states that the rate of fuel consumption is proportional to both the local limiting reactant concentration and the local rate of mixing. Extinction is based on the critical flame temperature approach~\cite{FDS_Math_Guide}. While FDS provides users with the option to use a more complex finite-rate combustion mechanism, there is not sufficient evidence in this case to justify deviating from the default specifications.

@@

Initial simulations were conducted using a HRR for the two fires totaling 5.3~MW, however, the fire behavior did not align with observations made during the fire incident or post-incident images. Specifically there was insufficient fuel to drive the basement to an underventilated condition. One reason is that while the structure was unoccupied at the time of the fire~\cite{PGCounty2013}, there were additional fuels that were not accounted for: transient fuels, sparse kitchen cabinetry, and an unknown bathroom vanity among other fuels. A second reason is that the 50~kW/m$^2$ HRRPUA used to estimate the fire size does not account for an observed increase in burning rate during wind driven fires~\cite{madrzykowski2009fire,kerber2009fire}. It is not clear how to quantify the impact of wind on burning rate within a structure from first principles, so several iterations of the model with increased fire sizes were run. The final estimated fire size was a 1.8~MW fire in the bathroom and a 6.8~MW fire in the kitchen. This resulted in a total peak specified fire size of 8.6~MW.

Since the time at which the fire started is not known relative to the notification of a fire or how fast the fire spread, several simplifications were made. The fires started at the beginning of the simulation and increased to a value of 8.6~MW in 10~s. The 10-s ramp up time was used to prevent abrupt changes in the velocity at the fire boundary condition in FDS. Since the actual ignition time is unknown, the fire was ramped quickly to reflect the conditions determined from reports and post-incident images. The second assumption is that the sources of the gas phase wood fuel have constant areas at fixed locations in the basement (1 source for the kitchen, 1 source for the bathroom).  At 207~s, using a simple decay model, the fire size decreases linearly to 25~\% of its peak value in 60~s. This is to account for the impact of firefighters putting water on the basement fire (Table~\ref{tab:fire_info}). The following lines define the source fires in the FDS input file:

@@

\section{Ventilation}
\label{Vents}
The simulation in this study accounts for changes in the ventilation due to a combination of fire department operations (opening doors), fire acting on the structure (fire breaching windows) and wind (restricting flow, closing doors). The ventilation times and ventilation areas represent the best understanding of the incident. The estimated times of ventilation changes for doors and windows are provided in Table~\ref{tab:vents}.

During the time when the firefighters exited the structure or were rescued from the structure, the state of some windows and doors in the structure was known. The state of the vents (opened versus closed) resulted in the establishment of a flow path in the interior stairwell leading from the basement to the first floor. The exact time of operation for other doors/windows was not known. Therefore, the following the two basement windows on the left rear corner of the structure were assumed to be open at the start of the simulation. The states of the front door and front window were specified at certain times based on observations (i.e. forcible entry, wind effects, forced window breakage) from post-incident reports.

Structure leakage has been shown to be important when modeling enclosure fires~\cite{beal2009}. The open rear basement windows at the start of the simulation were significantly larger than the typical total effective area of leakage of a structure of this type. Therefore, any leakage in the structure would have a negligible impact on the fluid mechanics within the structure and is not included in this study.

In addition to the state of openings within the structure, outlined in Table~\ref{tab:vents}, an 8.9~m/s (20~mph) wind velocity boundary condition was assumed. The flow direction of the wind was from the rear toward the front of the structure, perpendicular to the rear of the structure.  Based on local weather conditions, mean wind velocities ranged between 4.5~m/s (10~mph) and 8.9~m/s (20~mph) with gusts up to 13.4~m/s (30~mph) with a direction that angled toward the back left of the structure~\cite{PGCounty2013}.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{../Figures/pg_county_meshtotal2}
\caption[Overhead front-left side of the structure within the computational domain.]{Overhead front-left view of the structure within the computational domain (with 17 meshes). The entire computational domain was 24~m (79~ft) by 16~m (53~ft) by 8~m (26~ft).}
\label{fig:mult_mesh}
\end{figure}

@@

\begin{table}[!ht]
\centering
\captionof{table}{Relevant fire model input parameters.}\label{tab:mod_param}
\begin{tabular}{lll}
\toprule[1.5pt]
Parameter                                                  & Description                                   & Discussion                            \\
\midrule
Simulation Time                                            &  5~min                                        & --                                    \\ [.25cm]
\multirow{2}{*}{Grid Cell Size}                            &  10~cm around house                           & \multirow{2}{*}{Section~\ref{mesh}}   \\ 
                                                           &  20~cm for wind                               &                                       \\ [.25cm]
Ambient Temperature$^*$                                    &  20~$^{\circ}$C (68~$^{\circ}$F)              & --                                    \\ [.1cm]
\multirow{4}{*}{Reaction: Wood~\cite{SFPE:Tewarson}}       &  Formula: $\C\H_{1.7}\O_{0.74}\N_{0.002}$     & \multirow{4}{*}{Section~\ref{fire}}   \\
                                                           &  CO Yield: 0.004~$\mathrm{kg}/\mathrm{kg}$    &                                       \\
                                                           &  Soot Yield: 0.015~$\mathrm{kg}/\mathrm{kg}$  &                                       \\
                                                           &  Heat of Combustion: 16,400~kJ/kg             &                                       \\ [.25cm]
\multirow{2}{*}{Peak HRR}                                  &  Basement Fire: 6.75~MW                       & \multirow{2}{*}{Section~\ref{fire}}   \\ 
                                                           &  Bathroom Fire: 1.8~MW                        &                                       \\ [.25cm]                     
\multirow{3}{*}{Material: Wood~\cite{Incropera:1} }        &  $k$: 0.12~\si{W/(m.K)})                      & \multirow{3}{*}{Section~\ref{matl}}   \\
                                                           &  $\rho$: 510~kg/m$^3$                         &                                       \\
                                                           &  $c_{p}$: 1.38~\si{kJ/(kg.K)}                 &                                       \\ [.25cm]
\multirow{3}{*}{Material: Gypsum Board~\cite{WAKILI2007}}  &  $k$: 0.28~\si{W/(m.K)}                       & \multirow{3}{*}{Section~\ref{matl}}   \\ 
                                                           &  $\rho$: 810~kg/m$^3$                         &                                       \\ 
                                                           &  $c_{p}$: 1.0~\si{kJ/(kg.K)}                  &                                       \\
Wind 													   &  8.9~m/s (20~mph)							   & Section~\ref{Vents}                   \\
\bottomrule[1.25pt]
\end{tabular}\par
\footnotesize
$^{*}$Initial interior temperatures were assumed to be 20~$^{\circ}$C; exterior temperatures were 10~$^{\circ}$C~\cite{PGCounty2013}.
\normalsize
\end{table}

@@

\begin{table}[!ht]
\caption[Fire incident and simulation event timeline.]
{Fire incident and simulation event timeline.}
\begin{tabular}{ccl}
\toprule
Incident Time  &  Simulation  &  Fire Behavior / Fireground Operation                                \\
               &  Time        &                                                                      \\
{(hh:mm:ss)}   &  {(s)}       &                                                                      \\
\midrule
21:08:26       &              &  First 911 call is made.                                             \\
               &              &                                                                      \\
21:11:03       &              &  Dispatch for a``box'' alarm assignment- four engine companies       \\
               &              &  two truck companies, one rescue squad, and a commanding officer.    \\
               &              &                                                                      \\
21:21:20       &  0           &  FDS simulation begins.                                              \\
               &              &                                                                      \\
21:12:55       &              &  First unit (Engine 807B) arrived on-scene.                          \\
               &              &                                                                      \\
21:13:44       &              &  Volunteer Chief 809A establishes incident command (IC).             \\
               &              &                                                                      \\
21:13:56       &              &  Truck 809 arrived on-scene.                                         \\
               &              &                                                                      \\
21:14:00       &  100         &  Front door opened.                                                  \\
               &              &                                                                      \\
21:14:42       &              &  Truck 809 enters structure.                                         \\
               &              &                                                                      \\
21:15:47       &  207         &  Front door closed over.                                             \\
               &              &  Basement door opened and water applied to fire.                     \\
               &              &                                                                      \\
21:15:51       &  211         &  Lower pane of front egress window broken.                           \\
               &              &                                                                      \\
21:16:01       &  221         &  Top pane of front egress window broken.                             \\
               &              &                                                                      \\
21:16:16       &              &  Truck 809 officer bailout through front window.                     \\
               &              &                                                                      \\
21:16:47       &  267         &  Front door opened.                                                  \\
               &              &                                                                      \\
21:17:15       &              &  Truck 809 forcible entry firefighter rescued.                       \\
               &              &                                                                      \\
\bottomrule
\end{tabular}
\footnotesize
\\ For specific information regarding ventilation areas used in the simulation, refer to Table~\ref{tab:vents}.
\normalsize
\label{tab:combined_timeline}
\end{table}

@@

\section{Heat Release Rate}
\label{HRR}
In FDS, a HRR must be specified, which results in a specified amount of fuel vapor (or pyrolyzate) being released from a fuel surface. FDS then determines the amount of combustion that occurs throughout the simulation based on the amount of available fuel and oxygen at a given location. As a result, the prescribed HRR that was input into the simulation can be different than the HRR calculated by FDS based on the ventilation conditions.

Figure~\ref{fig:PG_Total_HRR} shows the HRR in the structure vs. time based on the fires described in Section~\ref{fire}. In this figure, the solid line represents the prescribed HRR that was input into the simulation (based on a prescribed mass flux), and the dashed line represents the HRR inside the structure that was calculated by FDS based on the ventilation conditions. The vertical lines represent the times at which ventilation occurred: front door opening and closing as well as the breaking of a front window for emergency escape. See Table~\ref{tab:vents} for a detailed list of the ventilation that occurred during the incident. The maximum size of the fire specified in this study was selected such that the conditions aligned with observations and fuel loads from the actual fire incident.

\begin{figure}[!ht]
\includegraphics[width=5in]{../Figures/PG_Total_9MW_HRR}
\caption[Interior prescribed and calculated interior HRRs vs. time from the simulation.]
{Comparison of prescribed and calculated interior HRRs from the FDS simulation. The vertical lines indicate the times when ventilation occurred.}
\label{fig:PG_Total_HRR}
\end{figure}

In Fig.~\ref{fig:PG_Total_HRR}, the prescribed HRR that was input into the simulation is different than the HRR calculated by FDS based on the ventilation conditions. From 0~s to approximately 9~s, the calculated HRR was in approximate agreement with the prescribed HRR because there was an adequate amount of oxygen for all of the fuel to combust. The differences in prescribed versus predicted HRR is a result of either fuel accumulation within the structure not combusting because of a lack of oxygen or combustion occurring exterior to the structure where fuel and oxygen can mix and react. Fig.~\ref{fig:PG_9MW_HRR} shows the total calculated HRR within the computational domain (includes combustion occurring exterior to the structure) compared to the prescribed HRR curve.

@@

A comparison of the differences between the HRR inside the structure (Fig.~\ref{fig:PG_Total_HRR}) and the total HRR (Fig.~\ref{fig:PG_9MW_HRR}) over the first 100 seconds of the simulation indicated that on average there was the equivalent of an approximate 0.7 MW fire(s) occurring through open external vents (basement windows). This aligns with observed post-fire damage and post-incident report~\cite{PGCounty2013}. Fig.~\ref{fig:ext_fire_dam} shows a post-fire image of the back left corner of the structure as well as a Smokeview rendering of the fire from the FDS simulation prior to the front door being opened by firefighters.

@@

After 100 seconds of simulation time, 5~\sfrac{1}{2}~minutes after the first 911 call, firefighters open the front door to make entry to the first floor of the structure. The change in ventilation to the structure impacted the HRR. After the door was opened combustion no longer tool place outside of the structure, as shown in Fig.~\ref{fig:PG_Basement_9MW_HRR}, however the basement HRR was lower than the total HRR within the domain.

\begin{figure}[!ht]
\includegraphics[width=5in]{../Figures/PG_Basement_9MW_HRR}
\caption[Prescribed, calculated domain, and calculated basement HRRs vs. time from the simulation.]
{Comparison of prescribed, total calculated domain HRR and basement calculated HRRs from the simulation. The vertical lines indicate the times when ventilation changed.}
\label{fig:PG_Basement_9MW_HRR}
\end{figure}
